---
title: "Project1.Rmd"
author: "Karl Snyder"
date: "April 25, 2016"
output: 
  html_document:
    keep_md: true
---

# Project 1 from the "Reproducable Research" course.

## Load required packages and data.

This assumes the datafile is in a subfolder off the working directory

```{r}
library(ggplot2);library(plyr)
activity <- read.csv("./activity/activity.csv")  
```

## What is the mean and median total number of steps taken per day?

Use aggregate to get total steps by day, calculate the mean and median values, then create a histogram.

```{r}
byday <- with(activity, aggregate(steps, by = list(date), FUN = "sum", na.rm = TRUE))
  x <- byday$x
    avgstep <- mean(x, na.rm = TRUE)
    medstep <- median(x, na.rm = TRUE)
  print("Average");avgstep
  print("Median"); medstep

plot.new()
hist(x,main = "Histogram of Daily Steps",xlab = "Steps", ylab = "Days", col=3)
abline(v=avgstep, col="red",lty = 3)
text(x=avgstep,y=25,label="Average",col="red",cex=0.6)
text(x=avgstep,y=22,label=as.integer(avgstep),col="red",cex=0.6)
```

## What is the average daily activity pattern?

Use aggregate to get total steps by Interval, determine the interval with with highest average steps, then create a line graph.

```{r}
byint <- with(activity, aggregate(steps, by = list(interval),
                                  FUN = "mean", na.rm = TRUE))
    names(byint) <- c("interval","steps")
mstep <- max(byint$steps)
mint <- byint[byint$steps==mstep,1]
print("Interval for Max Averages Steps"); mint

plot.new()
plot(x=byint$interval, y=byint$steps, type = "l",
     main = "Average Steps by Interval",ylab = "Steps",xlab = "Interval",
     ylim = c(0,mstep*1.2))
abline(v=mint, col="red",lty = 3)
text(x=mint+100,y=mstep,label=mint,col="red")
```

## Imputing missing values

Calculate the number of missing values

```{r}
nastep <- sum(is.na(activity$steps))
print('Missing values');nastep
```

Replace missing values with interval means, recalculate mean and median, then replot the histogram

```{r}
act2 <- join(activity, byint, by="interval", type = "left", match = "all")
names(act2) <- c("steps","date","interval","avg_steps")
act2 <- transform(act2, steps= ifelse(is.na(steps),avg_steps,steps))

byday2 <- with(act2, aggregate(steps, by = list(date), FUN = "sum", na.rm = TRUE))
    x2 <- byday2$x
    avgstep2 <- mean(x2)
    medstep2 <- median(x2)
print("Average");avgstep2
print("Median"); medstep2

plot.new()
par(mfrow=c(1,2), mar=c(4,4,2,1), oma = c(0,0,2,0))
{ hist(x,main = "Exclude Missing Values",xlab = "Steps", ylab = "Days", col=3)
    abline(v=avgstep, col="red",lty = 3)
    text(x=avgstep,y=25,label="Average",col="red",cex=0.6)
    text(x=avgstep,y=22,label=as.integer(avgstep),col="red",cex=0.6)
  hist(x2,main = "With Imputed Values",xlab = "Steps", ylab = "Days", col=3)
    abline(v=avgstep2, col="red",lty = 3)
    text(x=avgstep2,y=30,label="Average",col="red",cex=0.6)
    text(x=avgstep2,y=27,label=as.integer(avgstep2),col="red",cex=0.6)
  mtext("Histograms of Daily Steps", outer = TRUE, cex = 2.0)}
```

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor "weekdays" or "weekends", then create a panel plot that compares activity patterns.

``` {r}
act2$date <- as.Date(act2$date)
act2 <- mutate(act2, day = weekdays(act2$date, abbreviate = TRUE))
act2 <- mutate(act2, wkday = ifelse(day == "Fri" | day == "Sat",
                                       "Weekend","Weekday"))

byint2 <- with(act2, aggregate(steps, by = list(interval,wkday),
                              FUN = "mean", na.rm = TRUE))
names(byint2) <- c("interval","wkday","steps")

plot.new()
p <- ggplot(byint2,aes(interval,steps))
p +
  geom_point(aes(shape = wkday, color = wkday))+
  geom_line(aes(linetype = wkday, color = wkday))+
  facet_grid(wkday~.)+
  labs( title = "Comparison of Weekday vs Weekend Activity",
        x = "Interval",
        y = "Avg Steps")+
  theme(legend.position="bottom")
```
